---
title: "Using BATL for SRM peak identification"
subtitle: "BCH 8110 Basecamp"
author: "Justin Chitpin"
date: "November 23, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/home/le/github_repo/BATL/")
```


## Downloading and installing everything required to use BATL

### Install packages required to install BATL

BATL uses functions from packages written by some other R developers. For BATL
to install correctly, these packages must be installed prior to BATL. The
following lines of code check if the 7 listed packages need to be installed on
your machine, and installs them for you if they are missing.

```{r, echo = TRUE, warning = FALSE, message = FALSE, error = FALSE}
## Install packages required for BATL if not already installed
packages_for_batl <- c("data.table", "doFuture", "future", "foreach", "igraph", "openxlsx", "progressr")
download_packages <- packages_for_batl[!(packages_for_batl %in% installed.packages()[, "Package"])]
if (length(download_packages) != 0) {
    install.packages(download_packages, repos = "http://cran.us.r-project.org")
}
```

Don't worry if you get the error:
`## Error in readRDS(file) : error reading from connection`.
This appears to be an error with the function `installed.packages()`

### Names of downloaded files for this workshop

These variables store the name of the BATL package, the BATL model file, and the
SRM peak file that you will be labelling. This is often better than manually
typing a long string of characters. Furthermore, if you need to change the SRM
file, for example, you can just change the variable name below.

```{r}
batl_package <- "batl_2.05.2.tar.gz"
batl_model <- "model_sphingolipid_v10.txt"
srm_file <- "BCH8110_20200909_COVID19_Lipophenomics_SL_Longitudinal_JC.txt"
lexicon <- "LipidLexiconforBasecamp2020.xlsx"
```

The following lines of code contain if statements. If the condition written in
the parentheses is true, then the code indented inside is evaluated. These if
statements check if our files of interest are contained in our working
directory. 

```{r}
if (!(batl_package %in% list.files())) {
    stop(paste(batl_package, "is not present in the working directory."))
}
if (!(batl_model %in% list.files())) {
    stop(paste(batl_model, "is not present in the working directory."))
}
if (!(batl_model %in% list.files())) {
    stop(paste(batl_model, "is not present in the working directory."))
}
if (!(lexicon %in% list.files())) {
    stop(paste(lexicon, "is not present in the working directory."))
}
```

### Installing BATL package

Installing the BATL package contained in our working directory if BATL has not
been installed already.

```{r}
if (batl_package %in% list.files()) {
    to_download <- "batl"[!("batl" %in% installed.packages()[, "Package"])]
    if (length(to_download) != 0) {
        install.packages(batl_package, repos = NULL, type = "source", dependencies = TRUE)
    }
}
```

## Introduction to using BATL

As a reminder, it is one thing to download an R package and it is another thing
to load it in order to use these package functions. The `library()` command
loads the package; double quotes around the package name are optional.

```{r, message = FALSE, warning = FALSE}
## Load libraries
library("batl")
library(data.table) # also works! (this is called non-standard evaluation btw)
library(openxlsx)
```

### Inspecting the model file

The following BATL function opens the BATL model file. The variable
`imported_model` is actually a list containing 2 objects. The first one is a
table containing the model information. The second one is a vector of meta
information for the model. 

```{r}
imported_model <- nb_import_model(model_name = batl_model)
```


You can use the `View()` command to open these variables up in RStudio as a
spreadsheet. Using `View()` is often easier than switching between RStudio and
your spreadsheets in Excel.

```{r, eval = FALSE}
View(imported_model$Model) # contains our lipid identities and model parameters
View(imported_model$Meta_information) # contains some relevant info
```

### Inspecting the SRM peak file

To view our SRM peak file, first load it into R as a variable. The peak file can
be imported into R using the `fread()` function which is part of the
`data.table` package. The first row of the file is treated as the column names
and `sep = "\t"` tells the function to expect tab-delimited values.

```{r}
imported_srm <- fread(srm_file, header = TRUE, sep = "\t")
```

Once again, you can use the `View()` command to open the peak file in RStudio.
Use the left and right arrows at the top of the panel to move across the
columns. This is one of the downsides of viewing very wide spreadsheets in R.

```{r, eval = FALSE}
View(imported_srm)
```

### Opening up the BATL documentation

Enter the following lines in the console to open up the documentation for the
BATL package. Vignettes are detailed, minimal-working examples that teach the
user how to use the package functions on sample data. You can replace `batl`
with the name of any other package to look up that package documentation. This
is very handy when learning how to use new functions.

```{r, eval = FALSE}
## Opening up documentation and vignette for the BATL package
help(package = "batl")
vignette(package = "batl", topic = "batl-Introduction")
```

### Using BATL

These next lines of code might seem strange but they are all of the parameters
we need to run the subsequent BATL functions.

```{r}
## Save name of the SRM peak file once we've labelled it
savenames <- paste0(gsub(".txt","", srm_file), "_labelled.txt")

## Variables for the artifact/isotope annotating function
lipid_category <- "SP"   # enable sphingolipid-specific artifact annotations
retention_tolerance <- 2 # number of significant figures after decimal place
machine_tolerance <- 0.5 # Q1 m/z tolerance

## Variables for the peak labelling function
qstandard <- "C16-D31 Ceramide" # internal standard
qstandard_col <- "Component Name" # column name where the qstandard is located
subtract_constant <- 100 # if subtracted RT is used; same value as training set
exceptions <- c("ADC", "TIC") # rows to avoid barcoding based on Mass Info column

```

The `deartifact_peaks` function loads the srm peak file. Lipids within the
sphingolipid category $SP$ may dehydrate (once or twice), deglycosylate (one or
more times), or dimerize in-source. These artifacts do not occur for other types
of lipids lacking hydroxyl groups or 6-carbon sugars. Isotopes, on the other
hand, may be present for any type of lipid. The `exceptions` parameter is
required because some rows in the SRM peak file are not actually SRM peaks! Rows
under the Mass.Info column containing the keywords TIC (total ion chromatogram)
and ADC (analog digital converter) are sometimes output by the peak picking
software MultiQuant. Because these rows do not reflect SRM peaks, they are
excluded from the annotation function.

```{r}
deartifacted_files <- deartifact_peaks(
    filenames = srm_file,
    category = lipid_category,
    Q1_tolerance = machine_tolerance,
    retention_tolerance = retention_tolerance,
    exceptions = exceptions)
```

Note how the variable returned is a list. This is because `srm_file` may be a
vector of SRM peak files. As we only ran a single SRM file into the function,
the variable returned is a list containing a single table. We can view the table
and check any artifact or isotope annotations in the last column called
`Insource_annotation`.

```{r, eval = FALSE}
View(deartifacted_files[[1]])
```

As you can see, the output of this function is the exact same SRM peak file that
we input. The only difference is the additional column called
`Insource_annotation`.  

Most of the rows under the `Insource_annotation` column are blank or NA. Instead
of manually scrolling through the table in RStudio, we can use the following
line of code to subset the table to only include rows with isotopes and
artifacts.

```{r, eval = FALSE}
View(deartifacted_files[[1]][!(is.na(Insource_annotation))])
```

Because the peak labelling function inputs the name of the SRM peak file, we
need to temporarily export the newly annotated SRM peak file. The following for
loop will export all of SRM peak files using the `savenames` variable we created
above. In our case, we only have one file to export but this loop is handy when
exporting multiple files at once.

```{r}
for (i in seq_along(deartifacted_files)) {
    deartifact_export_peaks(
        file = deartifacted_files[[i]], filename = savenames[i])
}
```

The `nb_label_peaks` function assigns lipid labels to our SRM peak file. Some of
the features such as Area, Height, Relative.RT, and Subtracted.RT are normalized
to the corresponding feature value of the internal standard within the same
sample. For that reason, we need to specify `qstandard` and `qstandard_col`.
`subtract_constant` is necessary to avoid negative values because the likelihood
cannot be estimated for the log-normal distributed (the logarithm of a negative
number is undefined).

```{r}
labelled_files <- nb_label_peaks(
    filenames = savenames, # the srm peak file with insource annotations
    model_filename = batl_model, # BATL model file
    qstandard = qstandard, # internal standard
    qstandard_col = qstandard_col, # column name where the internal standard is found
    subtract_constant = subtract_constant, # for the subtracted RT variable
    exceptions = exceptions)
```

The following for loop will export all of SRM peak files using the `savenames`
variable we created above. In our case, we only have one file to export but this
loop is handy when exporting multiple files at once.

```{r}
## Export
for (i in seq_along(labelled_files)) {
    nb_export_labelled_file(
        labelled_file = labelled_files[[i]], filename = savenames[i])
}
```

You could either view the labelled SRM peak file in RStudio or in Excel.
Note how the last column `Barcode_MWBM` contains the assigned lipid identities
for each peak. A row with no peak or a row with a peak that has no matching
lipid identities in the training set will be assigned an NA/blank barcode entry.
A peak that matched one or more lipids in the training set based on transition
but was not assigned a lipid identity is assigned the `UNASSIGNED` barcode.

```{r, eval = FALSE}
View(labelled_files[[1]])
```

#### Exercise 1

The "openxlsx" package allows you to open and export xlsx spreadsheets. Using
the help() function, can you figure out which openxlsx function writes a table
in R to an xlsx spreadsheet?

#### Exercise 2
I've stated that the `deartifact_peaks` and `nb__label_peaks` can accept a
vector of multiple SRM peak files. Can you modify the code above to deartifact
and label multiple SRM peak files? (Hint: although you only have a single SRM
peak file, you could create a vector of that file.)

### "VLOOKUP" in R

Each barcode under the `Barcode_MWBM` column represents a unique lipid identity.
These barcode-lipid identities are stored in the Excel spreadsheet saved in the
`lexicon` variable. While you could use VLOOKUP in Excel to merge these lipid
identities to the SRM peak file, you could also do this very easily in R.

```{r}
lexicon_table <- read.xlsx(xlsxFile = lexicon)
lexicon_table <- lexicon_table[, c("Barcode", "MolecularID")]
```

The `merge()` function is the simplest function to merge two tables in R. The
merging is based on the column names specified in `by.x` and `by.y`. In other
words, if the same Barcode is found in both tables, the MolecularID value in
table `y` will be appended to the corresponding row in table `x`. The 
`all.x = TRUE` parameter will append an NA value if there is no matching
Barcode between both tables.

```{r}
merged_table <- merge(
    x = labelled_files[[1]],
    y = lexicon_table,
    by.x = "Barcode_MWBM",
    by.y = "Barcode",
    all.x = TRUE)
```

Exporting the labelled SRM peak file for the last time!

```{r}
nb_export_labelled_file(labelled_file = merged_table, filename = savenames)
```

